Type Error Telemetry at Scale

### 2023-09-25 programming23/paper.tex
conclusion

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\percentile}[1]{{#1}th~percentile}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\nK}[1]{#1\textsc{k}}
\newcommand{\stddev}[1]{[\nK{#1}]}
\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1\vphantom{Ap}}}
\newcommand{\obox}[1]{\colorbox{orange!29}{#1\vphantom{Ap}}}
\newcommand{\rbox}[1]{\colorbox{red!30}{#1\vphantom{Ap}}}
\newcommand{\tefsfont}[1]{\textsc{#1}}
\newcommand{\tekey}{\tefsfont{te}}
\newcommand{\fskey}{\tefsfont{fs}}
\newcommand{\QALAN}{\FILL{} Alan:}
\newcommand{\panon}{pseudonymized}

@@

In this paper, we investigate methods for measuring the effectiveness
of the {Luau} type system in the development of {Roblox} scripts.
We want to collect measurements at a large scale, with thousands
of participants, and we must do so in a way that protects creators' privacy.
In comparison to prior work~(\cref{s:related}), which
with few exceptions~\cite{zhlbr-cc-2020,zhlbr-oopsla-2020,hlzbr-ecoop-2021} is either small in scale
or collects personally identifiable information~(PII),
we performed a large-scale study using \panon{} \emph{telemetry}.

@@

@@

\section{Lessons Learned}
\label{s:conclusion}

Quickly summarize findings:

\begin{itemize}
  \item
    \pct{50} of errors in current module, so probably errors appear locally (not in a distant module);
    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    \pct{90} \mnocheck{},  \pct{10} \mnonstrict{},  \pct{0.1} \mstrict{}, wow.
  \item
  \item
  \item
    DM awareness is an important and surprising gap between strict
    and forced strict, undoubtedly a usability barrier for \mstrict{}.
  \item
\end{itemize}

Quick regrets, lessons learned, advice for next time

\paragraph{Future work: Low adaption}

why so few using nonstrict and strict?
needs advertising?
needs usability?

\paragraph{Do we need timestamps?}

FILL these were helpful for the overview

Our telemetry records include timestamps.
While timestamps are not personal data per se, they are in the ballpark.
Telemetry would be even more private without them.
Do we need them?

Timestamps are not the focus of our analysis, but they do play a critical role.
Consider~\cref{f:error-density}.
With timestamps, it is clear that error density stays close to zero.
Without timestamps, we would be left with the counts at the bottom
of~\cref{f:error-density}.
Many distributions could fit those counts, such as a ``mountain'' session that first
grows and grows the number of type errors then shrinks and shrinks down to zero.

\acks

>>>>

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FILL}{\textbf{FILL}}
\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
\newcommand{\chkYes}{\dotscale{\CIRCLE}}
\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
\newcommand{\chkNo}{\wideas{}{\chkYes}}
\newcommand{\pct}[1]{\SI{#1}{\percent}}
\newcommand{\percentile}[1]{{#1}th~percentile}
\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}
\newcommand{\FS}{background}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\nK}[1]{#1\textsc{k}}
\newcommand{\stddev}[1]{[\nK{#1}]}
\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1\vphantom{Ap}}}
\newcommand{\obox}[1]{\colorbox{orange!29}{#1\vphantom{Ap}}}
\newcommand{\rbox}[1]{\colorbox{red!30}{#1\vphantom{Ap}}}
\newcommand{\tefsfont}[1]{\textsc{#1}}
\newcommand{\tekey}{\tefsfont{te}}
\newcommand{\fskey}{\tefsfont{fs}}
\newcommand{\QALAN}{\FILL{} Alan:}
\newcommand{\panon}{pseudonymized}

@@

%% TODO migratory

In this paper, we investigate methods for measuring the effectiveness
of the {Luau} type system in the development of {Roblox} scripts.
We want to collect measurements at a large scale, with thousands
of participants, and we must do so in a way that protects creators' privacy.
In comparison to prior work~(\cref{s:related}), which
with few exceptions~\cite{zhlbr-cc-2020,zhlbr-oopsla-2020,hlzbr-ecoop-2021} is either small in scale
or collects personally identifiable information~(PII),
%% TODO careful, proprietary matters too!!!
we performed a large-scale study using \panon{} \emph{telemetry}.

@@

% TODO
% \begin{table}
%   \caption{Modes and analyses}
%   \label{t:mode-analyze}
% 
%   Mode: nocheck ...
%   Analysis: nocheck ... forced strict
% 
% \end{table}

@@

\section{Reflections and Lessons Learned}
\label{s:conclusion}

This paper presents the first large-scale analysis of developers' interactions with
an industrial strength typechecker.
With +1.5 million telemetry records from over three hundred thousand sessions
that took place during a three-month timespan, the data has plenty to say about
type errors and their overlap with program edits.
At the same time, the data is carefully restricted to avoid leaking personal
and proprietary information: we know the number and kind of type errors
that occurred, but nothing about the line(s) of code that the typechecker
complained about.

On the object level, the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all creators (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    this mode exists.
    \QALAN{} nonstrict requires no annotations, right?
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Creators tend to fix them immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, the \mstrict{} mode clearly needs to relax its data-asset
    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
\end{itemize}

On the meta level, the main takeaway is that the lightweight, counter-based telemetry
in Roblox Studio supports a variety of useful analyses:
\begin{itemize}
  \item
    Counting specific type errors helps to identify trends and to confirm that
    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
  \item
    Testing the relationship between two counts can yield insights;
    in~\cref{f:tefs-compass}, the gap between \mstrict{} vs \FS{} errors
    revealed an issue with \mstrict{} checking.
  \item
    Reporting the overlap between type errors and the current (approximate) edit
    range lets us assess the outcome of edits without revealing any source code.
    It was especially useful to know the overlap for old and for current type errors.
  \item
    Timestamps give useful, low-level insights about the frequency of keystrokes,
    the length of sessions, and actions over time~(\cref{f:indy-session}).
    The event ordering that timestamps provide was invaluable; e.g., for learning
    that error deltas bounce up and down~(\cref{f:error-density}).
\end{itemize}

One aspect of the telemetry that we would change for next time is to provide
more metadata.
Luau recently adopted semantic subtyping to reduce false-positive type
errors~\cite{luau-subt}, but there is no way to tell if this
internal extension is enabled without an extra metadata field.
Two other aspects are worth rethinking.
First, it would be useful to have fine-grained counts for \FS{} analysis
errors and to know the specific reason behind errors such as \code{TypeMismatch}.
But, this additional data could double the size of telemetry records.
Second, tracking edit ranges made the telemetry system much more difficult to
build and maintain.
It would be much easier to track old and current errors at the module level,
but it remains to be seen how coarse of an approximation that would be.

In future work, it would be interesting to build statistical models of
the ``average'' programmer using each typing mode and compare their error
rates.
It would be possible to start modeling with our \mnocheck{} and \mnonstrict{}
data, but we have too few \mstrict{} records at the moment due to our uniform
sampling rate.
More data is needed  for a robust model of \mstrict{} programmers.

%% PS downloading from Kibana was a HUGE pain and source of errors. Awful!

%% -----------------------------------------------------------------------------
\acks
